---
title: "Picrust"
author: "IY"
date: "2026-03-23"
output: html_document
---

# ggpicrust2 Error Bar Script

This script was adapted from "R Scripts/Module 15 - ggpicrust2.Rmd," originally written by Avril Metcalfe-Roach

```{r library, include=FALSE}
library(tidyverse)
library(phyloseq)
library(ggpicrust2)
```
# Picrust Analysis of KO Term Data
This pipeline will be a complex analysis of the Picrust output we did on the class server.

## Input and Convert the Needed Files

Input and convert the ```phyloseq_taxonomy.rds``` sample data into a data frame and move the ```fastq``` sample IDs into a column. Then import the ```pred_metagenome_unstrat.tsv``` file.

```{r Import Data}
meta = readRDS('../datasets/ps_filtered.rds') %>% .@sam_data %>% data.frame() %>% 
  rownames_to_column('sample_name')

ko = read_delim('../datasets/picrust_output/KO_metagenome_out/pred_metagenome_unstrat.tsv')
```

## Group

```{r Group}
fast = c('accelerator','crusier sprinter','manoeuvrer')
meta = meta %>%
  mutate(swim_group = ifelse(swim_performance %in% fast,
                      'Fast',
                      ifelse(swim_performance == 'generalist',
                      'Moderate',
                      'Slow'
                      )
  ),
  .after = swim_performance
  )
meta_fast_slow = meta %>% 
  filter(
    swim_group == c('Fast','Slow')
  )
```


## Run the Pipeline

```{r DAA}
# Filter to only compare hindgut microbiome
hindgut = meta_fast_slow %>% filter(sample_type %in% c('hindgut'))
ko_filt = select(ko,`function`) %>% 
  mutate(`function` = sapply(.$`function`, function(x) str_split(x,'ko:') %>%
                            .[[1]] %>%
                            .[2])
    )

ko_joined = add_column(ko, ko_filt$`function`)

# Rename the ko value so the first column is rownames
ko_named = column_to_rownames(ko_joined, var = 'ko_filt$`function`')
ko_selected = select(ko_named, -`function`)

ko_daa = ko_selected [
  ,hindgut$sample_name
]

# sapply(ko$`function, function(x) str_remove(x,'ko:') %>% unlist())

# Perform pathway differential abundance analysis (DAA) using LinDA method
daa_results_df = pathway_daa(abundance = ko_daa,
                             metadata = hindgut, 
                             group = "swim_group", 
                             daa_method = "LinDA", 
                             select = NULL, reference = NULL)

saveRDS(daa_results_df, '../datasets/fish_daa_results_df.rds')

daa_results_df = readRDS('../datasets/fish_daa_results_df.rds')
```

## Convert KO to KEGG

```{r KO to KEGG}
# Annotate pathway results using KO to KEGG conversion
# This will take a while to run as it downloads KEGG data - save as an .rds file if you want to avoid re-running it
daa_annotated_results_df = pathway_annotation(pathway = "KO",
                                              daa_results_df = daa_results_df,
                                              ko_to_kegg = TRUE)

saveRDS(daa_annotated_results_df,'../datasets/fish_daa_annotated_results_df.rds')

daa_annotated_results_df<-readRDS('../datasets/fish_daa_annotated_results_df.rds')
```

## This script ends here

There was no significant pathways in the annotated results. In light of this, we created a volcano plot to highlight the non-significance.